---
title: "Hiearchical Ice off model"
output: pdf_document
---

```{r, message = FALSE}
library(tidyverse)
library(lubridate)
library(caret)
library(recipes)
library(readr)
library(rstan)
library(ggplot2)
library(reshape2)

data_ip <- read_csv("data_ip.csv")
lake_data <- read_csv("ltbl_ice.csv")
```

```{r}
merged_data <- left_join(data_ip, lake_data, by = "station_id")

merged_data <- merged_data %>%
  mutate(
    year_ice_off = year(ice_off),
    ice_off_doy = yday(ice_off),
    # Add 365 to DOY if ice_off happens in the year after the lake season year
    ice_off_doy = if_else(year_ice_off < year, ice_off_doy - 365, ice_off_doy)
  ) %>%
  filter(!is.na(ice_off))  # Only keep records with ice-off data
```

```{r}
features <- c("year", "lat_wgs84", "altitude_m", "area_ha", "country")
target <- "ice_off_doy"

rec <- recipe(as.formula(paste(target, "~", paste(features, collapse = "+"))), data = merged_data) %>%
  step_log(area_ha, base = exp(1)) %>%
  
  step_range(year, min = -1, max = 1) %>%
  
  step_normalize(all_numeric_predictors(), -year) %>%
  
  step_dummy(country, one_hot = TRUE)
```


```{r, message=FALSE}
rec_time <- recipe(ice_off_doy ~ year, data = merged_data) %>%
  step_range(year, min = -1, max = 1)  # scale year to [-1, 1]

rec_prep <- prep(rec_time, training = merged_data)
data <- bake(rec_prep, new_data = merged_data)

data$year2 <- data$year^2

stan_data_simple <- list(
  N = nrow(data),
  year = data$year,
  year2 = data$year2,
  y = data$ice_off_doy
)

suppress_output <- function(expr) {
  invisible(capture.output(result <- eval(expr, envir = parent.frame())))
  result
}
```

```{r, message=FALSE}
library(rstan)
library(ggplot2)
library(reshape2)
library(recipes)
library(dplyr)

rstan::rstan_options(auto_write = TRUE)
options(mc.cores = 1)

rec_prep <- prep(rec, training = merged_data)
data <- bake(rec_prep, new_data = merged_data)

country_vars <- grep("^country_", names(data), value = TRUE)

stan_data_extended <- list(
  N = nrow(data),
  C = length(country_vars),
  year = data$year,
  year2 = data$year^2,
  log_area = data$area_ha,
  altitude = data$altitude_m,
  latitude = data$lat_wgs84,
  country = as.matrix(data[, country_vars]),
  y = data$ice_off_doy
)

suppress_output <- function(expr) {
  invisible(capture.output(result <- eval(expr, envir = parent.frame())))
  result
}
```

### 4). **Fully Hierarchical Time Trend Model (Intercept, Linear, and Quadratic Effects)**

In this model, we take the hierarchical approach one step further by allowing **all three time-related parameters** to vary across lakes:

- The **intercept** \( \beta_{0i} \)
- The **linear time coefficient** \( \beta_{1i} \)
- The **quadratic time coefficient** \( \beta_{2i} \)

Each of these parameters is modeled as a function of lake-specific environmental features (such as log-transformed area, altitude, latitude, and country), plus random variation.

#### Model structure:

$$
y_i \sim \mathcal{N}(\mu_i, \sigma^2)
$$

$$
\mu_i = \beta_{0i} + \beta_{1i} \cdot \text{year}_i + \beta_{2i} \cdot \text{year}_i^2
$$

Where:

- \( \beta_{0i} \sim \mathcal{N}(\mu^{(0)}_i, \sigma_{\beta_0}^2) \)
- \( \beta_{1i} \sim \mathcal{N}(\mu^{(1)}_i, \sigma_{\beta_1}^2) \)
- \( \beta_{2i} \sim \mathcal{N}(\mu^{(2)}_i, \sigma_{\beta_2}^2) \)

Each mean \( \mu^{(k)}_i \) (for \( k = 0, 1, 2 \)) is defined by environmental covariates:

#### Intercept:
$$
\mu^{(0)}_i = w_0 + w_{\text{area}} \cdot \log(\text{area}_i) + w_{\text{alt}} \cdot \text{altitude}_i + w_{\text{lat}} \cdot \text{latitude}_i + \sum_c \gamma^{(0)}_c \cdot \text{country}_{ic}
$$

#### Linear time effect:
$$
\mu^{(1)}_i = v_0 + v_{\text{area}} \cdot \log(\text{area}_i) + v_{\text{alt}} \cdot \text{altitude}_i + v_{\text{lat}} \cdot \text{latitude}_i + \sum_c \gamma^{(1)}_c \cdot \text{country}_{ic}
$$

#### Quadratic time effect:
$$
\mu^{(2)}_i = q_0 + q_{\text{area}} \cdot \log(\text{area}_i) + q_{\text{alt}} \cdot \text{altitude}_i + q_{\text{lat}} \cdot \text{latitude}_i + \sum_c \gamma^{(2)}_c \cdot \text{country}_{ic}
$$

This model allows each lake to have its own:
- Baseline ice-on day (\( \beta_{0i} \))
- Linear trend over time (\( \beta_{1i} \))
- Nonlinear (accelerating/decelerating) trend over time (\( \beta_{2i} \))

All while sharing information across lakes through environmental structure. This offers maximum flexibility for capturing heterogeneous climate responses across space and time.

```{r}
stan_model_code_hier3 <- "
data {
  int<lower=1> N;
  int<lower=1> C;
  vector[N] year;
  vector[N] year2;
  vector[N] log_area;
  vector[N] altitude;
  vector[N] latitude;
  matrix[N, C] country;
  vector[N] y;
}

parameters {
  // Intercept model
  real w0;
  real w_area;
  real w_altitude;
  real w_latitude;
  vector[C] w_country;
  real<lower=0> sigma_beta0;
  vector[N] beta0_raw;

  // Linear slope model
  real v0;
  real v_area;
  real v_altitude;
  real v_latitude;
  vector[C] v_country;
  real<lower=0> sigma_beta1;
  vector[N] beta1_raw;

  // Quadratic slope model
  real q0;
  real q_area;
  real q_altitude;
  real q_latitude;
  vector[C] q_country;
  real<lower=0> sigma_beta2;
  vector[N] beta2_raw;

  // Residual SD
  real<lower=0> sigma;
}

transformed parameters {
  vector[N] mu_env_0;     // mean for intercept
  vector[N] mu_env_1;     // mean for linear slope
  vector[N] mu_env_2;     // mean for quadratic slope

  vector[N] beta0;
  vector[N] beta1;
  vector[N] beta2;
  vector[N] mu;

  for (i in 1:N) {
    mu_env_0[i] = w0
                + w_area * log_area[i]
                + w_altitude * altitude[i]
                + w_latitude * latitude[i]
                + dot_product(w_country, country[i]);

    mu_env_1[i] = v0
                + v_area * log_area[i]
                + v_altitude * altitude[i]
                + v_latitude * latitude[i]
                + dot_product(v_country, country[i]);

    mu_env_2[i] = q0
                + q_area * log_area[i]
                + q_altitude * altitude[i]
                + q_latitude * latitude[i]
                + dot_product(q_country, country[i]);

    beta0[i] = mu_env_0[i] + beta0_raw[i] * sigma_beta0;
    beta1[i] = mu_env_1[i] + beta1_raw[i] * sigma_beta1;
    beta2[i] = mu_env_2[i] + beta2_raw[i] * sigma_beta2;

    mu[i] = beta0[i] + beta1[i] * year[i] + beta2[i] * year2[i];
  }
}

model {
  // Priors for intercept regression
  w0 ~ normal(0, 5);
  w_area ~ normal(0, 2);
  w_altitude ~ normal(0, 2);
  w_latitude ~ normal(0, 2);
  w_country ~ normal(0, 2);
  sigma_beta0 ~ normal(0, 2);
  beta0_raw ~ normal(0, 1);

  // Priors for linear slope regression
  v0 ~ normal(0, 5);
  v_area ~ normal(0, 2);
  v_altitude ~ normal(0, 2);
  v_latitude ~ normal(0, 2);
  v_country ~ normal(0, 2);
  sigma_beta1 ~ normal(0, 2);
  beta1_raw ~ normal(0, 1);

  // Priors for quadratic slope regression
  q0 ~ normal(0, 5);
  q_area ~ normal(0, 2);
  q_altitude ~ normal(0, 2);
  q_latitude ~ normal(0, 2);
  q_country ~ normal(0, 2);
  sigma_beta2 ~ normal(0, 2);
  beta2_raw ~ normal(0, 1);

  // Residual error
  sigma ~ cauchy(0, 2.5);

  // Likelihood
  y ~ normal(mu, sigma);
}

generated quantities {
  vector[N] log_lik;
  for (i in 1:N) {
    log_lik[i] = normal_lpdf(y[i] | mu[i], sigma);
  }
}
"
```

```{r, eval=FALSE}
stan_model_hier3 <- suppress_output(
  stan_model(stanc_ret = stanc(model_code = stan_model_code_hier3))
)

fit_hier3 <- suppress_output(
  sampling(
    stan_model_hier3,
    data = stan_data_extended,
    iter = 2000,
    chains = 1,
    seed = 123,
    refresh = 500
  )
)

saveRDS(fit_hier3, "fit_hier3_ice_off.rds")
```

```{r}
fit_hier3 <- readRDS("fit_hier3_ice_off.rds")

# Get the actual country abbreviations from the data
country_vars <- grep("^country_", names(data), value = TRUE)
country_abbreviations <- gsub("country_", "", country_vars)

# Prepare the parameter names using actual country abbreviations
w_country_params <- paste0("w_country[", 1:length(country_abbreviations), "]")
v_country_params <- paste0("v_country[", 1:length(country_abbreviations), "]")
q_country_params <- paste0("q_country[", 1:length(country_abbreviations), "]")

# Create meaningful parameter names
w_country_named <- paste0("w_country[", country_abbreviations, "]")
v_country_named <- paste0("v_country[", country_abbreviations, "]")
q_country_named <- paste0("q_country[", country_abbreviations, "]")

# Combine all parameters into one list for the summary
summary_params <- c(
  # Intercept regression
  "w0", "w_area", "w_altitude", "w_latitude",
  # Linear slope regression
  "v0", "v_area", "v_altitude", "v_latitude",
  # Quadratic slope regression
  "q0", "q_area", "q_altitude", "q_latitude",
  # Variance terms
  "sigma", "sigma_beta0", "sigma_beta1", "sigma_beta2",
  # Country effects with proper names
  w_country_params, v_country_params, q_country_params
)

# Print the summary with the updated parameter names
summary_output <- summary(fit_hier3, pars = summary_params)$summary
rownames(summary_output)[grep("w_country\\[", rownames(summary_output))] <- w_country_named
rownames(summary_output)[grep("v_country\\[", rownames(summary_output))] <- v_country_named
rownames(summary_output)[grep("q_country\\[", rownames(summary_output))] <- q_country_named

print(summary_output)

# Traceplot with meaningful labels
traceplot(fit_hier3, pars = summary_params, inc_warmup = FALSE)
```

```{r, echo=FALSE}
# Extract posterior samples
samples_hier3 <- rstan::extract(fit_hier3, pars = c(
  # Global predictors
  "w0", "w_area", "w_altitude", "w_latitude",
  "v0", "v_area", "v_altitude", "v_latitude",
  "q0", "q_area", "q_altitude", "q_latitude",
  # Variance parameters
  "sigma", "sigma_beta0", "sigma_beta1", "sigma_beta2",
  # Country-level effects
  "w_country", "v_country", "q_country"
))

# Extract actual country abbreviations
country_vars <- grep("^country_", names(data), value = TRUE)
country_abbreviations <- gsub("country_", "", country_vars)

par(mfrow = c(4, 4), mar = c(4, 4, 2, 1))

# Plotting function
plot_posterior <- function(samples, name, xlab = name) {
  if (length(samples) > 1) {  # Check if samples exist
    plot(density(samples),
         main = paste("Posterior of", name),
         xlab = xlab,
         col = "blue", lwd = 2)
    abline(v = mean(samples), col = "red", lty = 2)
  } else {
    cat("No valid samples for:", name, "\n")
  }
}

# Plot global parameters
plot_posterior(samples_hier3$w0, "w0")
plot_posterior(samples_hier3$w_area, "w_area", "log(area)")
plot_posterior(samples_hier3$w_altitude, "w_altitude", "altitude")
plot_posterior(samples_hier3$w_latitude, "w_latitude", "latitude")

plot_posterior(samples_hier3$v0, "v0")
plot_posterior(samples_hier3$v_area, "v_area", "log(area)")
plot_posterior(samples_hier3$v_altitude, "v_altitude", "altitude")
plot_posterior(samples_hier3$v_latitude, "v_latitude", "latitude")

plot_posterior(samples_hier3$q0, "q0")
plot_posterior(samples_hier3$q_area, "q_area", "log(area)")
plot_posterior(samples_hier3$q_altitude, "q_altitude", "altitude")
plot_posterior(samples_hier3$q_latitude, "q_latitude", "latitude")

plot_posterior(samples_hier3$sigma, "sigma", "residual SD")
plot_posterior(samples_hier3$sigma_beta0, "sigma_beta0", "intercept SD")
plot_posterior(samples_hier3$sigma_beta1, "sigma_beta1", "linear slope SD")
plot_posterior(samples_hier3$sigma_beta2, "sigma_beta2", "quadratic slope SD")

# Plot country-specific parameters with proper abbreviations
num_countries <- length(country_abbreviations)

for (i in 1:num_countries) {
  country_name <- country_abbreviations[i]
  if (!is.null(samples_hier3$w_country)) {
    if (i <= ncol(samples_hier3$w_country)) {
      plot_posterior(samples_hier3$w_country[, i], paste0("w_country[", country_name, "]"))
    } else {
      cat("No data for:", paste0("w_country[", country_name, "]"), "\n")
    }
  }
}
for (i in 1:num_countries) {
  country_name <- country_abbreviations[i]
  if (!is.null(samples_hier3$v_country)) {
    if (i <= ncol(samples_hier3$v_country)) {
      plot_posterior(samples_hier3$v_country[, i], paste0("v_country[", country_name, "]"))
    } else {
      cat("No data for:", paste0("v_country[", country_name, "]"), "\n")
    }
  }
}
for (i in 1:num_countries) {
  country_name <- country_abbreviations[i]
  if (!is.null(samples_hier3$q_country)) {
    if (i <= ncol(samples_hier3$q_country)) {
      plot_posterior(samples_hier3$q_country[, i], paste0("q_country[", country_name, "]"))
    } else {
      cat("No data for:", paste0("q_country[", country_name, "]"), "\n")
    }
  }
}
```

```{r, echo=FALSE}
beta0_matrix <- rstan::extract(fit_hier3, pars = "beta0")$beta0
beta1_matrix <- rstan::extract(fit_hier3, pars = "beta1")$beta1
beta2_matrix <- rstan::extract(fit_hier3, pars = "beta2")$beta2

beta0_means <- apply(beta0_matrix, 2, mean)
beta1_means <- apply(beta1_matrix, 2, mean)
beta2_means <- apply(beta2_matrix, 2, mean)

par(mfrow = c(1, 3))

plot(density(beta0_means),
     main = "Smoothed intercepts",
     xlab = expression(beta[0][i]),
     col = "blue", lwd = 2)
abline(v = mean(beta0_means), col = "red", lty = 2)

plot(density(beta1_means),
     main = "Smoothed slopes",
     xlab = expression(beta[1][i]),
     col = "blue", lwd = 2)
abline(v = mean(beta1_means), col = "red", lty = 2)

plot(density(beta2_means),
     main = "Smoothed quadratic effects",
     xlab = expression(beta[2][i]),
     col = "blue", lwd = 2)
abline(v = mean(beta2_means), col = "red", lty = 2)
```

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

beta0_means <- apply(rstan::extract(fit_hier3, pars = "beta0")$beta0, 2, mean)
beta1_means <- apply(rstan::extract(fit_hier3, pars = "beta1")$beta1, 2, mean)
beta2_means <- apply(rstan::extract(fit_hier3, pars = "beta2")$beta2, 2, mean)

if ("station_id" %in% colnames(merged_data)) { #Using station_id as lake_id
  lake_ids <- unique(merged_data$station_id)
} else {
  stop("Error: 'station_id' column is missing from merged_data.  Cannot determine lakes.")
}

exploratory_df <- data.frame(
  lake_id = lake_ids,
  beta0 = 0,
  beta1 = 0,
  beta2 = 0,
  latitude = 0,
  altitude = 0,
  log_area = 0,
  country = ""
)

for (i in 1:length(lake_ids)) {
  current_lake_id <- lake_ids[i]
  
  lake_indices <- which(merged_data$station_id == current_lake_id)
  
  exploratory_df$beta0[i] <- beta0_means[lake_indices[1]]
  exploratory_df$beta1[i] <- beta1_means[lake_indices[1]]
  exploratory_df$beta2[i] <- beta2_means[lake_indices[1]]
  
  exploratory_df$latitude[i] <- mean(merged_data$lat_wgs84[lake_indices], na.rm = TRUE)
  exploratory_df$altitude[i] <- mean(merged_data$altitude_m[lake_indices], na.rm = TRUE)
  exploratory_df$log_area[i] <- mean(log(merged_data$area_ha[lake_indices]), na.rm = TRUE)
  
  exploratory_df$country[i] <- as.character(unique(merged_data$country[lake_indices])[1])
}

exploratory_long <- exploratory_df %>%
  pivot_longer(cols = c(beta0, beta1, beta2), names_to = "parameter", values_to = "value")

ggplot(exploratory_long, aes(x = latitude, y = value, color = factor(country))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, aes(color = factor(country))) + # Smooth lines by country
  facet_wrap(~ parameter, scales = "free_y", labeller = label_parsed) +
  labs(title = "Lake-specific parameters vs Latitude by Country",
       y = expression(beta[i]), x = "Latitude", color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot(exploratory_long, aes(x = altitude, y = value, color = factor(country))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, aes(color = factor(country))) + # Smooth lines by country
  facet_wrap(~ parameter, scales = "free_y", labeller = label_parsed) +
  labs(title = "Lake-specific parameters vs Altitude by Country",
       y = expression(beta[i]), x = "Altitude (m)", color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot(exploratory_long, aes(x = log_area, y = value, color = factor(country))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, aes(color = factor(country))) + # Smooth lines by country
  facet_wrap(~ parameter, scales = "free_y", labeller = label_parsed) +
  labs(title = "Lake-specific parameters vs Log Area by Country",
       y = expression(beta[i]), x = "log(Area)", color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

beta0_means <- apply(rstan::extract(fit_hier3, pars = "beta0")$beta0, 2, mean)
beta1_means <- apply(rstan::extract(fit_hier3, pars = "beta1")$beta1, 2, mean)
beta2_means <- apply(rstan::extract(fit_hier3, pars = "beta2")$beta2, 2, mean)

if ("station_id" %in% colnames(merged_data)) { #Using station_id as lake_id
  lake_ids <- unique(merged_data$station_id)
} else {
  stop("Error: 'station_id' column is missing from merged_data.  Cannot determine lakes.")
}

country_df <- data.frame(
  lake_id = lake_ids,
  beta0 = 0,
  beta1 = 0,
  beta2 = 0,
  country = ""
)

for (i in 1:length(lake_ids)) {
  current_lake_id <- lake_ids[i]
  
  lake_indices <- which(merged_data$station_id == current_lake_id)
  
  country_df$beta0[i] <- beta0_means[lake_indices[1]]
  country_df$beta1[i] <- beta1_means[lake_indices[1]]
  country_df$beta2[i] <- beta2_means[lake_indices[1]]
  
  country_df$country[i] <- as.character(unique(merged_data$country[lake_indices])[1])
}

country_long <- country_df %>%
  pivot_longer(cols = c(beta0, beta1, beta2), names_to = "parameter", values_to = "value")

ggplot(country_long, aes(x = country, y = value, fill = country)) +
  geom_violin(alpha = 0.7, draw_quantiles = c(0.5), color = "black") +
  geom_jitter(width = 0.1, alpha = 0.04, size = 0.3, color = "black") +
  facet_wrap(~parameter, scales = "free_y", labeller = label_parsed) +
  labs(
    title = "Lake-specific β Parameters by Country",
    x = "Country",
    y = expression(beta[i])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  )
```

### Model Comparison:

```{r, message=FALSE, warning=FALSE}
library(loo)

log_lik_simple <- extract_log_lik(fit_hier3, merge_chains = FALSE)

loo_simple <- loo(log_lik_simple)
print(loo_simple)

cat("LOO Score for Simple Model:\n")
cat("elpd_loo: ", loo_simple$estimates["elpd_loo", "Estimate"], "\n")
cat("LOOIC: ", loo_simple$estimates["looic", "Estimate"], "\n")
```

```{r}
library(dplyr)
library(readr)

beta0_means <- apply(rstan::extract(fit_hier3, pars = "beta0")$beta0, 2, mean)
beta1_means <- apply(rstan::extract(fit_hier3, pars = "beta1")$beta1, 2, mean)
beta2_means <- apply(rstan::extract(fit_hier3, pars = "beta2")$beta2, 2, mean)

if ("country" %in% colnames(merged_data)) {
  country_info <- merged_data$country
} else {
  country_info <- "Unknown"
}

export_df <- data.frame(
  station_id = merged_data$station_id, # Taking station_id directly from merged data
  year = merged_data$year,             # Including year for each measurement
  beta0 = beta0_means,
  beta1 = beta1_means,
  beta2 = beta2_means,
  latitude = data$lat_wgs84,
  altitude = data$altitude_m,
  log_area = log(data$area_ha),
  country = country_info
)

write_csv(export_df, "lake_parameters_per_measurement.csv")

cat("Export complete: lake_parameters_per_measurement.csv")
```